Full Title

### 2018-01-28 cones/paper/cones.tex
Wrote section about Tsukada-Ong games.

\begin{corollary}
  The composition of strategies $\sigma\from A\implies B$ and $\tau\from B \implies C$ is a well-formed strategy for $A\implies C$.
  \label{cor:diverge-or-reply}
\end{corollary}
\begin{proof}
  The only tricky point is eastablishing that diverge-or-reply holds for $\sigma;\tau$.  
  Again, it is sufficient to prove this in the case that $\sigma$ and $\tau$ are deterministic and complete.
  Then it essentially follows from the argument used in \cite{abramskyjagadeesangames} that shows that a partiality at an $O$-position $s\in T_{\sigma;\tau}$ must arise either from a partiality in $T_\sigma$ or $T_\tau$ or from `infinite chattering' between $\sigma$ and $\tau$.  
  In the first case, the diverge-or-reply rule for $\sigma$ and $\tau$ gives us a divergence at $s$ in $\sigma;\tau$.  
  In the second case, an infinite chattering between $\sigma$ and $\tau$ corresponds to an infinite interaction $\s\in\int(A,B,C)$ ending with infinitely many moves in $B$ such that $\s\vert_{A,C}=s$. 
  Completeness for $\sigma$ and $\tau$ tells us that $\s\vert_{A,B}\in D_\sigma$ and $\s\vert_{B,C}\in D_\tau$ and therefore that $\s\vert_{A,C}\in D_{\sigma;\tau}$.  
\end{proof}

@@

In order to prove these, we come back to the concept of an \emph{evaluation} of a term as a sequence of natural numbers that we use to replace the constant $\wn$.

\begin{lemma}
  Let $M=C[\wn]$ be a term of type $\com$.  
  Write $\sigma_M$ for the denotation of the term $\lambda n.C[n]$.  
  Let $\pi$ be some evaluation of the configuration $\ia s M$, where $s$ is some state.  
  Then there is a deterministic total strategy $\sigma_\pi\from \oc\bN$ such that
  \begin{itemize}
    \item $\pi$ terminates at the value $\skipp$ if and only if then $qa\in T_{\sigma_\pi;\sigma_M}$.
    \item $\pi$ diverges if and only if $D_{\sigma_\pi;\sigma_M}\ne\emptyset$.
  \end{itemize}
  \label{lem:soundness-lemma}
\end{lemma}

\begin{proof}
  Let $n_1,\dots,n_k,d$ be a finite sequence of natural numbers.  
  We define an Idealized Algol term $N_{n_1,\dots,n_k,d}\from (\nat \to \com) \to \com$ to be the following.
  \[
    \lambda f.\new_\nat (\lambda v.f(v\coloneqq (suc\;\deref v); \case_{k+1}\;\deref v\;\Omega\;n_1 \cdots n_k d))\,.
    \]
  Here, $\case_{k+1}\;a\;n_0\;\cdots\;n_k\;d$ evaluates to $n_i$ if $a$ evaluates to $i$, and evaluates to $d$ if $a$ evaluates to $j> k$.
  Clearly, it can be built out of $\IfO$s.
  This term adds an extra variable $v$ to the program; each time $f$ is called, it increments the value of $v$ and maps its value on to one of the $n_i$.

@@

  Lastly, we need to deal with the case that $\pi$ is an evaluation $n_1,n_2,\dots$ that consults the nondeterministic oracle infinitely often.
  In this case, $M$ must certainly diverge along the evaluation $\pi$.
  In that case, we define $\pi_n^{(j)}$ to be the strategy for $\oc N$ corresponding to the term $N_{n_1,\dots,n_j,\Omega}$.  
  So $\pi_n^{(j)}$ has a unique finite maximal play
  \[
    q\;n_1\;q\;n_2\;\cdots\;q\;n_j\;q\,,
    \]
  at which point the strategy has a partiality.

@@

  $\sigma_\pi$ has plays of the form
  \[
    q\;n_1\;q\;n_2\;\cdots\,,
    \]
  and so it is total.
\end{proof}

@@

\begin{proof}[Proof of Lemma \ref{lem:consistency}]
  Suppose that that $M$ is a closed term of type $\com$ and that $M\converges\skipp$.  
  Let $\pi$ be an evaluation such that $M\converges\skipp$ along $\pi$.  
  Then $qa\in T_{\sigma_\pi;\sigma_M}$ for some strategy $\sigma_\pi$.  
  It follows that $qa\in T_{\top_\bN;\sigma_M}$.

  Now suppose that $D_{\deno M}\neq\emptyset$.  
  Then there is some interaction $\s\in\int_\infty(I,\oc\bN,\bC)$ such that $\s\vert_\bC\in D_{\sigma_M}$.
  Let $\sigma$ be the minimal deterministic strategy for $\oc\bN$ containing the play $\s\vert_{\oc\bN}$.  
  By the proof of Lemma \ref{lem:soundness-lemma}, $\sigma\subset\sigma_\pi$ for some evaluation $\pi$ of $M$.  
  Then we have $D_{\sigma_\pi;\sigma_M}\ne\emptyset$, so $M$ diverges along $\pi$.
\end{proof}

\begin{proof}[Proof of Proposition \ref{prop:adequacy}]
  Suppose that $M$ is a closed term of type $\com$ and that $qa\in T_{\deno M}$.  
  Then there is some interaction $\s\in\int_\infty(I,\oc\bN,\bC)$ such that $qa\in\s\vert_\bC$.  
  Let $\sigma$ be the minimal deterministic strategy for $\oc\bN$ containing the play $\s\vert_{\oc\bN}$.  
  By the proof of Lemma \ref{lem:soundness-lemma}, $\sigma\subset\sigma_\pi$ for some evaluation $\pi$ of the term $M$.
  Then we have $qa\in T_{\sigma_\pi;\sigma_M}$, so $M\converges\skipp$ along $\pi$.

  Now suppse that $M\not\mustconverge$.  
  Let $\pi$ be a divergent evaluation of $M$.  
  Then we have $D_{\sigma_\pi;\sigma_M}\ne\emptyset$.  
  Moreover, since $\sigma_\pi$ is total, it follows that $D_{\top_{\bN};\sigma_M}\ne\emptyset$.
\end{proof}

@@

>>>>

\begin{corollary}
  The composition of strategies $\sigma\from A\implies B$ and $\tau\from B \implies C$ is a well-formed strategy for $A\implies C$.
  \label{cor:diverge-or-reply}
\end{corollary}
\begin{proof}
  The only tricky point is eastablishing that diverge-or-reply holds for $\sigma;\tau$.  
  Again, it is sufficient to prove this in the case that $\sigma$ and $\tau$ are deterministic and complete.
  Then it essentially follows from the argument used in \cite{abramskyjagadeesangames} that shows that a partiality at an $O$-position $s\in T_{\sigma;\tau}$ must arise either from a partiality in $T_\sigma$ or $T_\tau$ or from `infinite chattering' between $\sigma$ and $\tau$.  
  In the first case, the diverge-or-reply rule for $\sigma$ and $\tau$ gives us a divergence at $s$ in $\sigma;\tau$.  
  In the second case, an infinite chattering between $\sigma$ and $\tau$ corresponds to an infinite interaction $\s\in\Int(A,B,C)$ ending with infinitely many moves in $B$ such that $\s\vert_{A,C}=s$. 
  Completeness for $\sigma$ and $\tau$ tells us that $\s\vert_{A,B}\in D_\sigma$ and $\s\vert_{B,C}\in D_\tau$ and therefore that $\s\vert_{A,C}\in D_{\sigma;\tau}$.  
\end{proof}

@@

Traditional proofs of computational adequacy using logical relations make essential use of the continuity of composition with respect to a natural ordering on strategies (see, for example, \cite{mcCHFiniteND} and \cite{RusssThesis} for the finite nondeterminism case).  
In our case, since composition is not continuous in the language itself, we cannot use this technique.
In order to prove soundness and adequacy, we use a new technique that involves using a deterministic stateful construction to model the nondeterminism inside a deterministic world in which continuity holds.  
We shall return to the concept of an \emph{evaluation} of a term as a sequence of natural numbers that we use to replace the constant $\wn$.

\begin{lemma}
  Let $M=C[\wn]$ be a term of type $\com$.  
  Write $\sigma_M$ for the denotation of the term $\lambda n.C[n]$.  
  \begin{itemize}
    \item $M\converges\skipp$ if and only if there exists some total deterministic strategy $\sigma\from\oc\bN$ such that $qa\in T_{\sigma;\sigma_M}$.
    \item $M\not{\mustconverge}$ if and only if there exists some total deterministic strategy $\sigma\from\oc\bN$ such that $D_{\sigma;\sigma_M}\ne\emptyset$.
  \end{itemize}
  \label{lem:soundness-lemma}
\end{lemma}

\begin{proof}
  We prove the forward directions first.

  Let $n_1,\dots,n_k,d$ be a finite sequence of natural numbers.  
  We define an Idealized Algol term $N_{n_1,\dots,n_k,d}\from (\nat \to \com) \to \com$ to be the following.
  \[
    \lambda f.\new_\nat (\lambda v.f(v\coloneqq (suc\;\deref v); \case_{k+1}\;\deref v\;\Omega\;n_1 \cdots n_k d))\,.
    \]
  Here, $\case_{k+1}\;a\;n_0\;\cdots\;n_k\;d$ evaluates to $n_i$ if $a$ evaluates to $i$, and evaluates to $d$ if $a$ evaluates to $j> k$.
  Clearly, it can be built out of $\IfO$s.
  This term adds an extra variable $v$ to the program; each time $f$ is called, it increments the value of $v$ and maps its value on to one of the $n_i$.

@@

  Lastly, we need to deal with the case that there is an infinite evaluation $\pi = n_1,n_2,\dots$ of $M$ that consults the nondeterministic oracle infinitely often.
  In this case, $M$ must certainly diverge along the evaluation $\pi$.
  For each $j$, we define $\pi_n^{(j)}$ to be the strategy for $\oc N$ corresponding to the term $N_{n_1,\dots,n_j,\Omega}$.  
  So $\pi_n^{(j)}$ has a unique finite maximal play
  \[
    q\;n_1\;q\;n_2\;\cdots\;q\;n_j\;q\,,
    \]
  at which point the strategy has a partiality.

@@

  $\sigma_\pi$ has plays of the form
  \[
    q\;n_1\;q\;n_2\;\cdots\,,
    \]
  and so it is total.

  We now prove the reverse implications.  
  Let $\sigma\from\bN$ be a total deterministic strategy for $\oc\bN$.
  Then $\sigma$ must have a maximal infinite play $s_\sigma$ of the form
  \[
    q\;m_1\;q\;n_2\;\cdots\,,
    \]
  where $m_1,m_2,\dots$ is some infinite sequence of natural numbers.  
  If the strategy $\sigma_M$ contains some play $\s$ such that $\s\vert_{\oc\bN}=s$, then $\sigma=\sigma_\pi$ for some infinite evaluation $\pi$ of $M$.  
  Otherwise, let $t$ be the maximal subplay of $s$ such that $\s\vert_{\oc\bN}=t$ for some $\s\in\sigma_M$.  
  Then, if we replace $\sigma$ with the strategy $\sigma'$ that plays according to $t$ and subsequently plays $q\;d\;q\;d\;\cdots$ for our fixed value $d$, then we will have $\sigma';\sigma_M=\sigma;\sigma_M$.  
  Moreover, $\sigma'=\sigma_\pi$ for some evaluation $\pi$ of $M$.

  Now suppose that there exists $\sigma\from\oc\bN$ such that $qa\in T_{\sigma;\sigma_M}$.  
  By the above discussion, we may assume that $\sigma=\sigma_\pi$ for some evaluation $\pi$ of $M$.  
  Therefore, $qa\in T_{\sigma_\pi;\sigma_M}$, which means that $M\converges\skipp$ along the evaluation $\pi$.  
  The corresponding statement for must convergence follows in the same way.
\end{proof}

@@

\begin{proof}[Proof of Lemma \ref{lem:consistency}]
  Suppose that that $M$ is a closed term of type $\com$ and that $M\converges\skipp$.  
  Then $qa\in T_{\sigma;\sigma_M}$ for some strategy $\sigma$.  
  It follows that $qa\in T_{\top_\bN;\sigma_M}$.

  Now suppose that $D_{\deno M}\neq\emptyset$.  
  Then there is some interaction $\s\in\Int_\infty(I,\oc\bN,\bC)$ such that $\s\vert_\bC\in D_{\sigma_M}$.
  Let $\sigma$ be a deterministic total strategy containing $\s\vert_{\oc\bN}$.  
  Then we have $D_{\sigma;\sigma_M}\ne\emptyset$, so $M$ diverges by Lemma \ref{lem:soundness-lemma}.
\end{proof}

\begin{proof}[Proof of Proposition \ref{prop:adequacy}]
  Suppose that $M$ is a closed term of type $\com$ and that $qa\in T_{\deno M}$.  
  Then there is some interaction $\s\in\Int_\infty(I,\oc\bN,\bC)$ such that $qa\in\s\vert_\bC$.  
  Let $\sigma$ be a deterministic total strategy for $\oc\bN$ containing the play $\s\vert_{\oc\bN}$.  
  Then we have $qa\in T_{\sigma;\sigma_M}$, so $M\converges\skipp$ by Lemma \ref{lem:soundness-lemma}.

  Now suppse that $M\not\mustconverge$.  
  Then we have $D_{\sigma;\sigma_M}\ne\emptyset$ for some strategy $\sigma\from\oc\bN$.
  Moreover, since $\sigma_\pi$ is total, it follows that $D_{\top_{\bN};\sigma_M}\ne\emptyset$.
\end{proof}

@@

\section{Nondeterministic PCF}

We conclude by saying a few remarks about the situation when our base deterministic language is PCF rather than Idealized Algol.  
Tsukada and Ong \cite{TsukadaSheaves} showed that traditional game semantics models are unsuitable for modelling this language: while the definition of \emph{visibility} of strategies can be extended to nondeterministic strategies without difficulty, the corresponding definition for PCF -- that of an \emph{innocent} strategy -- does not.  
Tsukada and Ong instead use a new type of game semantics in which strategies are given not by sets of plays but by justified \emph{trees} whose vertices are moves.  
The difference is subtle but important: from the point of view of countable nondeterminism it allows us to distinguish between terms that take an arbitrarily large finite number of steps to converge and terms that diverge.  
For example, the denotation of our term $M$ above has an infinite branch in it, while the denotation of $N$ has infinitely many finite branches of arbitrary length.

The main advantage of the tree-based model, however, is that it gives rise to a satisfactory synthetic definition of a \emph{nondeterministic innocent} strategy based on sheaves.  
In fact, Tsukada and Ong do not prove full abstraction for PCF with countable nondeterminism (even though that language may be interpreted within their category), but for PCF with finite nondeterminism.

Having ourselves given a fully abstract model of Idealized Algol with countable nondeterminism, we shall contribute to the process of `completing the square' by giving a sketch of how we can use our techniques to prove soundness and adequacy for the Tsukada-Ong model of PCF with countable nondeterminism.  
A side-effect of this will be to give an alternative proof of soundness and adequacy for their model of PCF with finite nondeterminism.

The first observation to make is that our Lemma \ref{lem:soundness-lemma} is not really a statement about the category $\G_{ND,vis}$, but one about the deterministic model $\G_{D,vis}$.

For example, the first part of the lemma says that if $C[-]$ is an ordinary deterministic Idealized Algol context, then $C[\wn]\converges\skipp$ if and only if there exists some total deterministic strategy $\sigma\from\oc\bN$ such that $qa\in T_{\sigma;\deno{\lambda n.C[n]}}$.  
Although we have mentioned a term ($C[\wn]$) living in nondeterministic Idealized Algol, this can be thought of as a statement about $C[-]$ itself, which is part of deterministic Idealized Algol.  
Moreover, the strategy $\sigma$ and the denotation of $\lambda n.C[n]$ are both deterministic strategies, so we can interpret Lemma \ref{lem:soundness-lemma} as telling us something about Idealized Algol and its game semantics.  
We can therefore forget about our nondeterminism model, with its divergence sets and infinite plays (neither of which are part of the Tsukada-Ong model) and concentrate on the Abramsky-McCusker model of Idealized Algol.
Note that although Lemma \ref{lem:soundness-lemma} mentions $T_{\sigma;\sigma_M}$ and $D_{\sigma;\sigma_M}$, we can rephrase both of these inside the deterministic model, since the strategies involved are both deterministic.

The second observation is that Lemma \ref{lem:soundness-lemma} can just as easily be applied to PCF contexts and its game semantics model $\G_{inn}$.  
There really is nothing to this observation: since any PCF context may be interpreted in Idealized Algol, then Lemma \ref{lem:soundness-lemma} applies to it.
So we have narrowed \ref{lem:soundness-lemma} down to a result about $\G_{inn}$.

Now we have an inclusion functor $\G_{inn}\hookrightarrow\G_{TO}$, where $\G_{TO}$ is the Tsukada-Ong category.  
Since the nondeterministic oracle $\wn$ may also be interpreted as a tree-strategy $\top_{\bN}$ in $\G_{TO}$, this gives us a model of PCF with countable nondeterminism inside that category.

It is then possible to deduce versions of Lemma \ref{lem:consistency} and Proposition \ref{prop:adequacy} from Lemma \ref{lem:soundness-lemma} as we did above, using the details of the inclusion functor from $\G_{inn}$ and the definition of composition in $\G_{TO}$.  
We can then deduce soundness in the usual way.

\begin{theorem}[Soundness for nondeterministic PCF]
  Let $M,N\from T$ be two terms of PCF with countable nondeterminism.
  If $\deno M\oes \deno N$, then $M\emam N$, where the denotation $\deno -$ is taken in the category $\G_{TO}$.  
\end{theorem}

An alternative line of attack is to first build a copy of our model of Idealized Algol with coutable nondeterminism inside $\G_{TO}$, using our argument to prove soundness and adequacy, and then restrict down to nondeterministic PCF.
